A lottery model of density-dependent selection in evolutionary genetics

### 2017-05-03 manuscript.tex
latest

\title{Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle}
%\author{Jason Bertram $^{1,\ast}$ \\ 
%Joanna Masel $^{1}$}

@@

In our representation of Grime's ideal environment, $d\ll 1$, $b_i/d_i \gg 1$ and $L\gg 1$, and so Eq. \eqref{eq:invad} becomes 
\begin{equation}
\Delta n_j/n_j \approx \frac{b_j}{c_j/c_i + L-1}\frac{c_j}{c_i}-d_j\approx \frac{b_j}{L}\frac{c_j}{c_i}-d_j, \label{eq:highLinvad}
\end{equation}
where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i$ (ignoring mutations with very large effect sizes; incidentally, Eq. \eqref{eq:lottery} yields the same expression by assuming $L\rightarrow \infty$ even though this assumption is unrealistic for a rare mutant). Similarly to Grime's stressful environment, improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits, even though they may have very different magnitudes, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium. Biases in the direction of evolution of $b$, $c$ or $d$ must again follow from biases in mutation availability or trade-offs/pleiotropy. The same argument applies for high-$L$ interpretation of the stress regime.

To represent Grime's disturbed environment, $\Delta_+ n_i$ is replaced with $(1-d_i)\Delta_+ n_i$, and a single type reaches equilibrium when $L/(1-e^{-L})=(1-d_i)b_i/d_i$ (population persistence now requires $(1-d_i)b_i>d_i$ rather than $b_i>d_i$). We assumed $d_i\approx 1$ and $(1-d_i)b_i\approx d_i$, so that $L\ll 1$. A mutant's invasion fitness is then given by
\begin{equation}
\Delta n_j/n_j = (1-d_j)b_j - d_j.
\end{equation}
Another distinct feature of disturbed environments is that the bulk of mortality is environmental and unavoidable. That is, $d_i=d_{ie}+d_{ig}$, where the environmental part $d_{ie}\approx 1$ does not evolve, and the genetic part $d_{ig}\ll 1$ is affected by mutations in the same way as before $d_{jg}=(1-\epsilon) d_{ig}$.

The effects of improving $b$ and $d$ are now no longer identical. Increasing $b$ by a fraction $\epsilon$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon$, whereas reducing $d$ by the same fraction gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$. Thus, even though the genetic component of mortality $d_{g}$ is small compared to the environmental component $d_{e}$, if disturbance is intense enough ($1-d_i$ is small enough), selection for reduced mortality is potentially much stronger than that for propagule production.

>>>>

\title{A lottery model of density-dependent selection in evolutionary genetics}
%\author{Jason Bertram $^{1,\ast}$ \\ 
%Joanna Masel $^{1}$}

@@

In our representation of Grime's ideal environment, $d\ll 1$, $b_i/d_i \gg 1$ and $L\gg 1$, and so Eq. \eqref{eq:invad} becomes 
\begin{equation}
\Delta n_j/n_j \approx \frac{b_j}{c_j/c_i + L-1}\frac{c_j}{c_i}-d_j\approx \frac{b_j}{L}\frac{c_j}{c_i}-d_j, \label{eq:highLinvad}
\end{equation}
where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i$ (ignoring mutations with very large effect sizes; incidentally, Eq. \eqref{eq:lottery} yields the same expression by assuming $L\rightarrow \infty$ even though this assumption is unrealistic for a rare mutant). Similarly to Grime's stressful environment, improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits, even though they may have very different magnitudes, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium. Biases in the direction of evolution of $b$, $c$ or $d$ must again follow from biases in mutation availability or trade-offs/pleiotropy. The same argument applies for the high-$L$ interpretation of the stress regime.

To represent Grime's disturbed environment, $\Delta_+ n_i$ is replaced with $(1-d_i)\Delta_+ n_i$, and a single type reaches equilibrium when $L/(1-e^{-L})=(1-d_i)b_i/d_i$ (population persistence now requires $(1-d_i)b_i>d_i$ rather than $b_i>d_i$). We assumed $d_i\approx 1$ and $(1-d_i)b_i\approx d_i$, so that $L\ll 1$. A mutant's invasion fitness is then given by
\begin{equation}
\Delta n_j/n_j \approx (1-d_j)b_j - d_j.
\end{equation}
Another distinct feature of disturbed environments is that the bulk of mortality is environmental and unavoidable. That is, $d_i=d_{ie}+d_{ig}$, where the environmental part $d_{ie}\approx 1$ does not evolve, and the genetic part $d_{ig}\ll 1$ is affected by mutations in the same way as before $d_{jg}=(1-\epsilon) d_{ig}$.

The effects of improving $b$ and $d$ are now no longer identical for given values of $d_i$ and $d_{ig}$. Increasing $b$ by a fraction $\epsilon$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i$, whereas reducing $d_g$ by the same fraction gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.

Not surprisingly, selection for lower $d$ is weaker than for greater $b$ if $d_{ig}$ is sufficiently small. However, assuming that there is some non-negligible genetic contribution to mortality

Thus, since $d_i\approx 1$, selection for lower $d$ is potentially much stronger than that for higher $b$.

even though the genetic component of mortality $d_{g}$ is small compared to the environmental component $d_{e}$, if disturbance is intense enough ( is small enough), .
